Query Optimization for Dynamic Imputation

### 2017-05-14 paper/experiments.tex
Update count fraction, planning times

\begin{figure}
\includegraphics[width=\columnwidth]{figures/running_times_combined_bar.png}
\caption{Runtimes for the queries in~\Cref{tbl:queries} using reference
    chained-equation decision trees algorithm. $\alpha=0$ (resp. $\alpha=1$) is
    \ProjectName{} run at the highest (resp. lowest) quality setting. Baseline
    is full table imputation followed by running the query. Each query and each
    value of $\alpha$ was run 200 times.}
\todo{update with new figure}   
\label{fig:runtimes}
\end{figure}

@@

\begin{table}
\centering
\input{tables/perf_summary.tex}
\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$ parameterizations, as compared to the baseline.
Queries optimized
    for quality ($\alpha=0$) generally achieve lower error than queries optimized for
    efficiency ($\alpha=1$). The count fraction column shows the number of tuples used in calculating each aggregate
     as a fraction of the number of tuples used when running the same query after imputing on the base table.
    A lower count share reflects more potential for errors.}
\todo{update with new figure}
\label{table:smape}
\end{table}

@@

\subsubsection{Baseline Runtime}
In many real-world cases, applying imputation to an entire dataset is prohibitively expensive.
For example, if we run CART imputation on the full ACS dataset, it completes in \acsbaseresultminutes{}.

In contrast, \ProjectName{} executes the following query:
\begin{lstlisting}[breaklines]
SELECT AVG(c0) FROM acs;
\end{lstlisting}
over the ACS dataset (also using CART imputation) in \acsimputedbzeroresult{} when optimizing for quality and \acsimputedboneresult{} when optimizing for runtime.
The performance difference comes from the fact that \ProjectName{} only needs to perform imputation on the single column required by the query.
Adding selection predicates would further reduce the query runtime by reducing the amount of required imputation.
An analyst could do the same, but tracking the required imputations would get increasingly complicated as the queries became more complex.
This highlights the benefit of using our system for early data exploration.

@@

\todo{Update: Planning times for our benchmark queries are reasonable, with a median planning time of \planningmediantime{} ms. over the benchmark queries, constituting \planningruntimepercent{}\% of total runtime.
In all cases, the optimizer returned a query plan within \planningmaxtime{} ms.}
The choice of $\alpha$ has no impact on planning times, because query selection is done after collecting the Pareto frontier.

>>>>

\begin{figure}
\includegraphics[width=\columnwidth]{figures/running_times_combined_bar.png}
\caption{Runtimes for the queries in~\Cref{tbl:queries} using reference
    chained-equation decision trees algorithm. $\alpha=0$ (resp. $\alpha=1$) is
    \ProjectName{} run at the highest (resp. lowest) quality setting. Baseline
    is full table imputation followed by running the query. Each query and each
    value of $\alpha$ was run 200 times.}
\label{fig:runtimes}
\end{figure}

@@

\begin{table}
\centering
\input{tables/perf_summary.tex}
\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$ parameterizations, as compared to the baseline.
Queries optimized
    for quality ($\alpha=0$) generally achieve lower error than queries optimized for
    efficiency ($\alpha=1$). The count fraction column shows the number of tuples used in calculating each aggregate
     as a fraction of the number of tuples used when running the same query after imputing on the base table.
    A lower count share reflects more potential for errors.}
\label{table:smape}
\end{table}

@@

\subsubsection{Baseline Runtime}
In many real-world cases, applying imputation to an entire dataset is prohibitively expensive.
For example, if we run chained-equation decision trees imputation on the full ACS dataset, it completes in \acsbaseresultminutes{}.

In contrast, \ProjectName{} executes the following query:
\begin{lstlisting}[breaklines]
SELECT AVG(c0) FROM acs;
\end{lstlisting}
over the ACS dataset (also using chained-equation decision trees imputation) in \acsimputedbzeroresult{} when optimizing for quality and \acsimputedboneresult{} when optimizing for runtime.
The performance difference comes from the fact that \ProjectName{} only needs to perform imputation on the single column required by the query.
Adding selection predicates would further reduce the query runtime by reducing the amount of required imputation.
An analyst could do the same, but tracking the required imputations would get increasingly complicated as the queries became more complex.
This highlights the benefit of using our system for early data exploration.

@@

\todo{Update: Planning times for our benchmark queries are reasonable, with mean planning
times across the benchmark queries ranging from \planninglowtime{} ms to
\planninghightime{} ms, constituting \planninglowtimepct{} to \planninghightimepct{} of
total runtime, respectively.

In all cases, the optimizer returned a query
plan within \planningmaxtime{} ms.} The choice of $\alpha$ has no impact on planning times,
because query selection is done after collecting the Pareto frontier.

### 2017-05-14 paper/main.tex
paper: mark sections for sam to review
Update count fraction, planning times

\documentclass{vldb}
\usepackage{balance}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[activate={true,nocompatibility},final,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
\usepackage[subtle]{savetrees}
\usepackage{amsfonts,amssymb,amsmath}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[style=numeric,firstinits=true]{biblatex}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{graphicx} % includegraphics
\usepackage{listings} % code samples with syntax hl
\usepackage{stmaryrd}
\usepackage{qtree} % for simple tree diagrams
\usepackage{caption}
\usepackage{subcaption}
\usepackage{xcolor}
\usepackage{hyperref} % load after other packages
\usepackage{tikz}
\usetikzlibrary{positioning}
\usepackage{cleveref}
\usepackage{xspace}
\usepackage[all]{nowidow}

@@

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}
\newcommand{\todo}[1]{{\color{red}\textbf{TODO:} #1}}
\newcommand{\todobox}[2]{{\color{red}\fbox{\parbox{\columnwidth}{\textbf{TODO:} \textit{#1}\\#2}}}}

% results that can be used in any section
\newcommand{\demorows}{10175}
\newcommand{\labexrows}{9813}
\newcommand{\acsbaseresultminutes}{355 minutes}
\newcommand{\acsbaseresulthours}{just under 6 hours}
\newcommand{\acsimputedbzeroresult}{4 seconds}
\newcommand{\acsimputedboneresult}{1 second}
\newcommand{\lowxalphazero}{10}
\newcommand{\lowsmapealphazero}{0}
\newcommand{\highxalphazeroexacs}{200}
\newcommand{\highxalphaoneexacs}{1000}
\newcommand{\highsmapealphaoneexacs}{24}
\newcommand{\highxalphaone}{16000}
\newcommand{\nullv}{NULL}
\newcommand{\planningruntimepercent}{1.8}
\newcommand{\planningmediantime}{1.5}
\newcommand{\planningmaxtime}{34}

@@

\begin{abstract}
  Missing values are common in data analysis and present a usability challenge.
  Users are forced to pick between removing tuples with missing values or creating a cleaned version of their data by applying a relatively expensive imputation strategy.
  Our system, \ProjectName{}, incorporates imputation into a cost-based query optimizer, performing necessary imputations on-the-fly for each query.
  This allows users to immediately explore their data, while the system picks the optimal placement of imputation operations.
  % This placement is based on the specific query issued and a user-provided parameter that trades off data quality and runtime performance.
  % We incorporate the imputation operators into the relational algebra, allowing for traditional optimizations to take place.
  We evaluate this approach on three real-world survey-based datasets.
  % from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}.
  Our experiments show that our query plans execute between \lowxalphazero{} and \highxalphazeroexacs{} times faster than first imputing the base tables.
  Furthermore, we show that the query results from on-the-fly imputation differ from the
  traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaoneexacs{}\%, in one error measure.
  Finally, we show that while dropping tuples with missing values that fail query constraints discards 3--88\% of the data, on-the-fly imputation loses only 0--10\%.
\end{abstract}

\input{introduction}
\input{examples}
\input{algorithm}
\input{experiments}
\input{relatedwork}
\input{conclusion}
\bibstyle{abbrv}
\balance
\printbibliography

>>>>

\documentclass[preprint]{vldb}
\usepackage{balance}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage[activate={true,nocompatibility},final,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
\usepackage[subtle]{savetrees}
\usepackage{amsfonts,amssymb,amsmath}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[style=numeric,firstinits=true]{biblatex}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{graphicx} % includegraphics
\usepackage{listings} % code samples with syntax hl
\usepackage{stmaryrd}
\usepackage{qtree} % for simple tree diagrams
\usepackage{caption}
\usepackage{subcaption}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usepackage{xspace}
\usepackage[all]{nowidow}
\usepackage{enumitem}
\usepackage{hyperref} % load after other packages
\usepackage{cleveref}

@@

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}
\newcommand{\todo}[1]{{\color{red}\textbf{TODO:} #1}}
\newcommand{\todobox}[2]{{\color{red}\fbox{\parbox{\columnwidth}{\textbf{TODO:} \textit{#1}\\#2}}}}
\newcommand{\review}[1]{{\color{blue}#1}}

% results that can be used in any section
\newcommand{\demorows}{10175}
\newcommand{\labexrows}{9813}
\newcommand{\acsbaseresultminutes}{355 minutes}
\newcommand{\acsbaseresulthours}{just under 6 hours}
\newcommand{\acsimputedbzeroresult}{4 seconds}
\newcommand{\acsimputedboneresult}{1 second}
\newcommand{\lowxalphazero}{10} % the smallest speed up, comparing a=0 to base table imputation
\newcommand{\highxalphazeroexacs}{140} % the largest speed up, comparing a=0 to base table imputation 
\newcommand{\highxalphaoneexacs}{1400} % the largest speed up, comparing a=0 OR a=1 to base table imputation 
\newcommand{\lowsmapealphazero}{0} % the smallest SMAPE difference, comparing a=0 to base table imputation
\newcommand{\highsmapealphazero}{6} % the largest SMAPE difference, comparing a=0 to base table imputation
\newcommand{\highsmapealphaoneexacs}{20} % the largest SMAPE difference, comparing a=1 to base table imputation
%\newcommand{\highxalphaone}{16000}
\newcommand{\nullv}{NULL}

\newcommand{\planninglowtime}{0.7}
\newcommand{\planninglowtimepct}{0}
\newcommand{\planninghightime}{2.8}
\newcommand{\planninghightimepct}{14}

\newcommand{\planningmaxtime}{34} % TODO

@@

\begin{abstract}
  Missing values are common in data analysis and present a usability challenge.
  Users are forced to pick between removing tuples with missing values or creating a cleaned version of their data by applying a relatively expensive imputation strategy.
  Our system, \ProjectName{}, incorporates imputation into a cost-based query optimizer, performing necessary imputations on-the-fly for each query.
  This allows users to immediately explore their data, while the system picks the optimal placement of imputation operations.
  % This placement is based on the specific query issued and a user-provided parameter that trades off data quality and runtime performance.
  % We incorporate the imputation operators into the relational algebra, allowing for traditional optimizations to take place.
  We evaluate this approach on three real-world survey-based datasets.
  % from the \textit{Center for Disease Control and Prevention} and \textit{freeCodeCamp}.
  Our experiments show that our query plans execute between \lowxalphazero{} and \highxalphazeroexacs{} times faster than first imputing the base tables.
  Furthermore, we show that the query results from on-the-fly imputation differ from the
  traditional base-table imputation approach by \lowsmapealphazero{}--\highsmapealphaoneexacs{}\%, in one error measure.
  Finally, we show that while dropping tuples with missing values that fail query constraints discards 6--78\% of the data, on-the-fly imputation loses only 0--21\%.
\end{abstract}

\input{introduction}
\input{examples}
\input{algorithm}
\input{implementation}
\input{experiments}
\input{relatedwork}
\input{conclusion}

\section{Acknowledgments}
The authors would like to acknowledge the authors of~\cite{akande2015empirical} for giving us a cleaned copy of the 2012 ACS PUMS.

\bibstyle{abbrv}
\balance
\printbibliography
